---
title: "Quick Peek into Oncology Data Science"
format: html
---

I was looking at the work of data scientists in Oncology and decided to give
a naive try at what they do. So, here it goes.

## Patient Journeys

An important caveat about understanding cancer is that treatment is not necessarily
about "curing" the disease; to summarise [Hank Green](https://www.youtube.com/watch?v=cepUz29tqEI),
of the 9000 Americans who develop Hodgkin Lymphoma, 900 die. "Boom, that's your statistic right
there, 10%, RIGHT? NOPE, WRONG!"

Survival rates are not designed to be interpreted as simply as that; according to the
CDC, such interpretations don't take into account factors like the age of the individual
at time of diagnosis, the duration of the disease, the stage of the cancer, and
sophistication of available treatment and technology, and many other factors.
You can see this from the following visualization:

```{r}
library(tidyverse)
library(ggrain)
library(plotly)
library(data.world)
```

```{r}
ds <- "https://data.world/makeovermonday/2018w40-five-year-cancer-survival-rates-in-america"

tables <- data.world::query(
    data.world::qry_sql("SELECT * FROM Tables"),
    dataset = ds
)

print(tables)
```

```{r}
query <- "SELECT * from five_year_cancer_survival_rates_in_usa"

five_yr <- data.world::query(
    data.world::qry_sql(query),
    dataset = ds
)

five_yr
```

From glimpsing this data, we can see that the term "survival rate"
can be broken down by race, gender, and cancer type. Essentially, each row
tells us what the 5-year survival rate would be that year for a person
who contracts that cancer. Let's look:

```{r}
five_yr %>% 
    ggplot(aes(x = 1, y = survival_rate, fill = race, colour = race)) +
    theme_minimal() +
    geom_rain(alpha = .5) +
    coord_flip() +
    theme(axis.text.y = element_blank(), axis.title.y = element_blank())+ 
    labs(title = "Survival rates of aspects of cancer over time", subtitle = "Race")
```

```{r}
five_yr %>%
     ggplot(aes(x = 1, y = survival_rate, fill = gender, colour = gender)) +
    theme_minimal() +
    geom_rain(alpha = .5) +
    coord_flip() +
    theme(axis.text.y = element_blank(), axis.title.y = element_blank())+ 
    labs(title = "Survival rates of aspects of cancer over time", subtitle = "Gender")
```

```{r}
five_yr %>%
    ggplot(aes(x = cancer_type, y = survival_rate)) +
    theme_minimal() +
    geom_violin() +
    theme(axis.text.x = element_text(angle = 45))+ 
    labs(title = "Survival rates of aspects of cancer over time", subtitle = "Cancer type")
```

These data appear to be a bit of a mish mosh, but remember that this is
_over time_. Let's try and put that in perspective by replicating the
[original plot](https://data.world/makeovermonday/2018w40-five-year-cancer-survival-rates-in-america).

Let's take a look at leukemia for example:

```{r}
five_yr %>%
    filter(cancer_type == "Leukemia") %>%
    ggplot(aes(x = year, y = survival_rate, colour = race)) +
    geom_line() +
    facet_grid(gender~cancer_type)
```

This shows an upward trend overall, but there wouldn't be enough reasonable room
to plot each and every combination of variables. We should reasonably compress some
of them. I'm going to assume that we wouldn't be missing out on too much
if we ignored intermittent downward trends, so maybe it would be interesting
 to see the change in survival rate over time?

```{r}
five_yr %>%
    filter(cancer_type == "Leukemia") %>%
    filter(!is.na(survival_rate)) %>%
    group_by(race, gender) %>%
    summarise(survival_rate_change = last(survival_rate) - first(survival_rate))
```

In this table, the percent of people who survive leukemia has increased
by the value in the rightmost column. We can think of this as an
"improvement in outcomes" measure over time. Let's do this for all cancers:

```{r}
five_yr %>%
    filter(!is.na(survival_rate)) %>%
    group_by(race, gender, cancer_type) %>%
    summarise(survival_rate_change = last(survival_rate) - first(survival_rate)) %>%
    arrange(survival_rate_change)
```

This is interesting. We can now see that improvement
varies by race, gender, and cancer type, with some survival
rates actually _decreasing_ over time. That's worrying. Let's plot
this to see the full gambit:

```{r}
five_yr %>%
    filter(!is.na(survival_rate)) %>%
    group_by(race, gender, cancer_type) %>%
    summarise(survival_rate_change = last(survival_rate) - first(survival_rate)) %>%
    arrange(survival_rate_change) -> five_yr_change

pp <- five_yr_change %>%
    ggplot(aes(y = survival_rate_change, x=cancer_type, fill=gender)) +
    geom_col(position = "dodge") +
    coord_flip() +
    theme_minimal() +
    facet_grid(~race) + 
    labs(title = "Change in % 5-year Survival Rates of Different Cancers from 1977 to 2013", subtitle = "Race",
    caption = "The plot shows how the percentage of people who survive 5 years after diagnosis has changed over the last 40 years.")

ggplotly(pp)
```

From this chart, we can conclude that there have been improvements in almost all
cancer treatments with a few exceptions — for example, skin cancer amongst black females
has seen a noticeable drop in 5-year survival rates.

This is a quick and dirty look at the data to get a sense of where we are, but overall
still speaks to the original comment — understanding survival rates is complicated
and all interpretations have to be looked at within their respective contexts.